---
title: "The projectable workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The projectable workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette steps through a typical `projectable` workflow, taking us from a 
simple dataframe via a metadata-rich table constructed using the `prj_tbl_*()` 
functions through to a projection of that table in `gt` format.

The `projectable` package was designed to gell with `gt` and the wider `tidyverse`,
so -- although it is by no means necessary -- we will use `dplyr` verbs throughout.

```{r setup}
library(projectable)
library(dplyr)
library(gt)
```

## Setting the table

The first step in the `projectable` workflow involves transforming an ordinary
dataframe into a metadata-rich table-like object that is fit for projection. We 
term this 'setting the table' and we perform it using the `prj_tbl_*()` functions
along with `col_*()` helpers.

Thus, to generate a frequency cross tabulation out of the `mtcars` dataset with
rows corresponding to the number of cylinders and the transmission types, and 
columns corresponding to the engine shape, we write:

```{r message = FALSE, warning = FALSE}
crs_tab <- mtcars %>% 
  # Specify the rows of the prj table
  prj_tbl_rows(
    Cylinders = cyl,
    Transmission = am,
  ) %>%
  # Specify the columns of the prj table
  prj_tbl_cols(
    `V-Shaped` = col_freq(n = vs %in% 1, N = vs %in% 0:1),
    `Not V-Shaped` = col_freq(n = vs %in% 0, N = vs %in% 0:1)
  ) %>%
  prj_tbl_summarise()

# Notice that the rows sum to one
knitr::kable(crs_tab)
```

Behind the scenes, `prj_tbl_summarise()` loops over the row and column expressions it is
provided. Each column expression provided via `prj_tbl_cols()` is evaluated over `.data` filtered
down to each row expression provided via `prj_tbl_rows()`.

In our example, `col_freq(...)` is evaluated over `mtcars` filtered to each 
unique value in `mtcars$cyl`, and then over `mtcars` filtered to each unique 
value in `mtcars$am`. 

There are helper functions which make specifying the columns slightly less verbose. 
Thus, an equivalent way to write the above is:

```{r}
crs_tab_simple <- mtcars %>% 
  # Specify the rows of the prj table
  prj_tbl_rows(
    Cylinders = cyl,
    Transmission = am,
  ) %>%
  # Specify the columns of the prj table
  prj_tbl_cols(!!!spec_col_freq(vs, 0:1, "row")) %>%
  prj_tbl_summarise()

# Notice that the rows sum to one
knitr::kable(crs_tab_simple)
```

If we wanted to calculate the base for each frequency, not row-by-row, but 
over the whole column we would have to explicitly instruct R, when it is evaluating
`N`, not to look for `vs` within the filtered version of `mtcars`, but in 
`mtcars` overall. Thus we would write

```{r}
crs_tab_col <- mtcars %>% 
  prj_tbl_rows(
    Cylinders = cyl,
    Transmission = am,
  ) %>% 
  prj_tbl_cols(
    `V-Shaped` = col_freq(n = vs %in% 1, N = .data$vs %in% 1), 
    `Not V-Shaped` = col_freq(n = vs %in% 0, N = .data$vs %in% 0)
  ) %>%
  # Equivalently: prj_tbl_cols(!!!spec_col_freq(vs, 0:1, "col")) %>%
  prj_tbl_summarise()

# Notice that, for each mutually exclusive and collectively exhaustive set of rows,
# the columns sum to one
knitr::kable(crs_tab_col)
```


And if we wanted to calculate the base for each frequency, not over the whole
column, but over the whole table, then we would write:

```{r}
# Calculate frequency cell-by-cell
crs_tab_overall <- mtcars %>% 
  prj_tbl_rows(
    Cylinders = cyl,
    Transmission = am,
  ) %>% 
  prj_tbl_cols(
    `V-Shaped` = col_freq(n = vs %in% 1, N = mtcars$vs %in% 0:1), 
    `Not V-Shaped` = col_freq(n = vs %in% 0, N = mtcars$vs %in% 0:1)
  ) %>%
  # Equivalently: prj_tbl_cols(!!!spec_col_freq(vs, 0:1, "cell")) %>%
  prj_tbl_summarise()

# Notice that, for each mutually exclusive and collectively exhaustive set of rows,
# the cells sum to one
knitr::kable(crs_tab_overall)
```


Once we have finished setting the table, what we should have is a table-like
object with `projectable_col`s. In addition to a face value, each of these 
`projectable_col`s will have a set of metadata associated with it: a `col_freq`, for 
instance, contains `n` and `N` metadata alongside a `p` (proportion) face value.

## Casting a shadow
When it comes to displaying this metadata-rich table-like object, we have a range
of options. 

In some circumstances, we may want to display only the face values of each 
`projectable_col`, in which case we can simply print the table as we have done above.

However, in many cases, we will also want to be able to display the associated 
metadata. Thus, there might be times when we want to display not only the 
proportions but the associated `n` and `N` too.

This is where the `prj_project()` function comes in: we use it to 'unfold' each
`projectable_col` into whichever of its constituent elements we would like to display. 
We do this by passing through a named list of glue-like character vectors where
each name corresponds to a column in the table passed through as `.data`. Each
element in the character vector will be pulled out into a column in the output.

So, for instance, to display the proportion alongside the count in one column 
and the base in another column for the `V-Shaped` and `Not V-shaped` columns we
created, we write:

```{r}
crs_tab_display <- crs_tab %>% 
  prj_project(
    .cols = list(
      `V-Shaped` = c(Frequency = "{signif(p, 2)} ({n})", Sample = "{N}"),
      `Not V-Shaped` = c(Frequency = "{signif(p, 2)} ({n})", Sample = "{N}")
    )
  )

knitr::kable(crs_tab_display)
``` 

We thus arrive at another 'flat' dataframe: a dataframe whose columns are simple
atomic vectors. But with one caveat, the dataframe as a whole comes associated
with metadata which keeps track of which columns in the output (what we call
the `projection`) belong to which columns in the input (the metadata rich
table). 

The significance of this metadata will become clear in the next section.  

## Shaping with `gt`
At this stage, the substance of what we want to display is already there. But 
its not in a form that's fit for general consumption. 

For one thing, while the columns in the `projection` give some indication of 
their provenance by way of their names, it would be good to make the connection
between input and output columns much more overt.

And for another, we might want to fiddle with the formatting: adding colours, 
touching up labels, annotating with footnotes. 

The bad news is that `projectable` can't do really any of this. The good news is 
that it doesn't need to. 

Instead of implementing any formatting functions, `projectable` implements a 
hand-off function, `prj_gt()`, which initialises a `gt` table object:

```{r}
crs_tab_gt <- crs_tab %>% 
  prj_shadow(where(is_col_freq), .shadow = c(Frequency = "{signif(p, 2)} ({n})", Sample = "{N}")) %>% 
  prj_gt()

crs_tab_gt
```

It is then possible to leverage all of the functionality of `gt` to apply 
whatever formatting we want:

```{r}
crs_tab_gt %>% 
  gt::tab_header(title = "Engine Shape vs Other Vehicle Characteristics") %>% 
  gt::tab_stubhead(label = "Vehicle Characteristics") %>% 
  gt::tab_options(
    heading.background.color = "#080808",
    row_group.background.color = "#f0f0f0"
  )
```

