---
title: "Programming with projectable"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Programming with projectable}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(projectable)
library(dplyr)
```


Other packages exist for producing summary tables in R. The main advantages
`projectable` is intended to have over these other packages are flexibility and
programmability.

Concretely, what that means is that the `projectable` user can specify *exactly*
what the columns and rows of the output table should include, and, moreover,
that they can do this programmatically.

Thus, taking the `mtcars` dataset as our example, if we wanted to see 
transmission type and engine shape broken down by a few different vehicle
characteristics, we could write:

```{r}
col_vars <- c("cyl", "gear")


my_tbls <- lapply(col_vars, function(cvar) {
  col_spec <- spec_col_freq(!!cvar, sort(unique(mtcars[[cvar]])))

  mtcars %>% 
    prj_tbl_cols(!!!col_spec) %>% 
    prj_tbl_rows(
      Transmission = am,
      `Engine Shape` = vs
    ) %>% 
    prj_tbl_summarise() %>% 
    prj_shadow_if(is_col(.), "{signif(p, 2)} ({n})") %>% 
    prj_project()
})

my_tbls
```

At the price of being a bit more verbose, we can gain finer-grained control
over the shape of the output. We can, for instance, omit automatic cars:

```{r}
val_labels <- list(
  cyl = c(`Four Cylinders` = 4, `Six Cylinders` = 6, `Eight Cylinders` = 8),
  gear = c(`Three Gears` = 3, `Four Gears` = 4, `Five Gears` = 5)
)

my_tbls <- mapply(col_vars, val_labels, FUN = function(cvar, cval) {
  col_spec <- spec_col_freq(!!cvar, cval)
  names(col_spec) <- names(cval)
  
  mtcars %>% 
    prj_tbl_cols(!!!col_spec) %>% 
    prj_tbl_rows(
      Transmission = list(Manual = am %in% 1),
      `Engine Shape` = list(
        `V-shaped` = vs %in% 1,
        `Not V-shaped` = vs %in% 0
      )
    ) %>% 
    prj_tbl_summarise() %>% 
    prj_shadow_if(is_col(.), "{signif(p, 2)} ({n})") %>% 
    prj_project()
}, SIMPLIFY = FALSE)

my_tbls
```

Or we can display only certain columns (this time doing column proportions 
rather than row proportions):

```{r}
val_labels <- list(
  cyl = c(`Four Cylinders` = 4, `Six Cylinders` = 6),
  gear = c(`Three Gears` = 3, `Four Gears` = 4)
)

my_tbls <- mapply(col_vars, val_labels, FUN = function(cvar, cval) {
  col_spec <- spec_col_freq(!!cvar, .vals = cval, "col", .base = unique(mtcars[[cvar]]))
  names(col_spec) <- names(cval)

  mtcars %>% 
    prj_tbl_cols(!!!col_spec) %>% 
    prj_tbl_rows(
      Transmission = am,
      `Engine Shape` = vs
    ) %>% 
    prj_tbl_summarise() %>% 
    prj_shadow_if(is_col(.), "{signif(p, 2)} ({n})") %>% 
    prj_project()
}, SIMPLIFY = FALSE)

my_tbls
```

